---
title: "Illustrating mvMAPIT with Simulated Data"
author: "Julian Stamp"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  Learn how to analyze data with mvMAPIT.
vignette: >
  %\VignetteIndexEntry{Illustrating mvMAPIT with Simulated Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.width=7,
  fig.height=5
)
library(knitr)
```

**_NOTE:_** if you use mvMAPIT in published research, please cite:

* Crawford, L., Zeng, P., Mukherjee, S., & Zhou, X. (2017). Detecting epistasis with the marginal epistasis test in genetic mapping studies of quantitative traits. PLoS genetics, 13(7), e1006869.
* Stamp, J., DenAdel, A., Weinreich, D. & Crawford, L. (2022). Multivariate MAPIT.
* Stamp J, Crawford L (2022). mvMAPIT: Multivariate Genome Wide Marginal Epistasis Test. https://github.com/lcrawlab/mvMAPIT, https://lcrawlab.github.io/mvMAPIT/


# Getting Started
Load necessary libraries. For the sake of getting started, `mvMAPIT` comes with
a small set of simulated data. This data contains random genotype-like data and two
simulated quantitative traits with epistatic interactions. To make use of this data,
call the genotype data as `simulated_data$genotype`, and the simulated trait data
as `simulated_data$trait`.
The vignette traces the analysis of simulated data. The simulations are described in `vignette("simulations")`.

For a working installation of mvMAPIT please look at the`README.md` or `vignette("docker-mvmapit")`
```{r load_dependencies, message=FALSE}
library(mvMAPIT)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyr)
library(kableExtra)
library(RcppAlgos)
```


# Running mvMAPIT

The R routine `mvmapit` can be run in multiple modes. By default it runs in a hybrid mode, performing tests both wtih a normal Z-test as well as the Davies method. The resulting `p`-values can be combined using functions provided by `mvMAPIT`, e.g. `fishers_combined()`, that work on the `pvalues` tibble that `mvmapit` returns.

>**NOTE:** mvMAPIT takes the X matrix as $p \times n$; **not** as  $n \times p$.
```{r run_mapit_normal, eval = FALSE}
mvmapit_hybrid <- mvmapit(
        t(data$genotype),
        t(data$trait),
        test = "hybrid"
)
fisher <- fishers_combined(mvmapit_hybrid$pvalues)
```


```{r assign_data, include = FALSE}
fisher <- mvmapit_data$fisher
pairs <- mvmapit_data$exhaustive_search
thresh <- 0.05 / (nrow(fisher) / 2)
```


To visualize the genome wide `p`-values, we use a Manhattan plot.
The `p`-values are plotted after combining the results from the multivariate analysis using Fisher's method.

```{r manhattan_plots}
manhplot <- ggplot(fisher, aes(x = 1:nrow(fisher), y = -log10(p))) +
  geom_hline(yintercept = -log10(thresh), color = "grey40", linetype = "dashed") +
  geom_point(alpha = 0.75, color = "grey50") +
  geom_text_repel(aes(label=ifelse(p < thresh, as.character(id), '')))
plot(manhplot)
```

To control the type I error despite multiple testing, we recommend the conservative Bonferroni correction.
The significant SNPs returned by the `mvMAPIT` analysis are shown in the output below. There are in total 6 significant SNPs after multiple test correction. Of the significant SNPs, 4 are true positives.

```{r significant_snps}
thresh <- 0.05 / (nrow(fisher) / 2)
significant_snps <-  fisher %>%
    filter(p < thresh) # Call only marginally significant SNPs

truth <- simulated_data$epistatic %>%
  ungroup() %>%
  mutate(discovered = (name %in% significant_snps$id)) %>%
  select(name, discovered) %>%
  unique()

significant_snps %>%
  mutate_if(is.numeric, ~(as.character(signif(., 3)))) %>%
  mutate(true_pos = id %in% truth$name) %>%
  kable(., linesep = "") %>%
  kable_material(c("striped"))
```

## True Epistatic SNPs

To compare this list to the full list of causal epistatic SNPs of the simulations, refer to the following list. There are 5 causal SNPs. Of these 5 causal SNPs, 4 were succesfully discovered by mvMAPIT.

```{r simulated_snps}
truth %>%
  kable(., linesep = "") %>%
  kable_material(c("striped"))
```

# Running an Informed Exhaustive Search

Now we may take only the significant SNPs according to their marginal epistatic effects and run a simple exhaustive search between them.

The search itself is a simple regression on the interaction terms between all significant interactions.


```{r search_significant_SNPs, eval = FALSE}
# exhaustive search for p-values
pairs <- NULL
if (nrow(significant_snps) > 1) {
  pairnames <- comboGeneral(significant_snps$id, 2)
  # Generate unique pairs of SNP names;
  # for length(names) = n, the result is a (n * (n-1)) x 2 matrix with one row corresponding to a pair
  for (k in seq_len(nrow(pairnames))) {
    fit <- lm(y ~ X[, pairnames[k, 1]]:X[, pairnames[k, 2]])
    p_value1 <- coefficients(summary(fit))[[1]][2, 4]
    p_value2 <- coefficients(summary(fit))[[2]][2, 4]
    tib <- dplyr::tibble(
            x = p_value1,
            y = p_value2,
            u = pairnames[k, 1],
            v = pairnames[k, 2]
    )
    pairs <- bind_rows(pairs, tib)
  }
}

colnames(pairs) <- c(colnames(y), "var1", "var2")
```

## Visualize Exhaustive Search Results

We plot the $-\mathrm{log}_{10}(p)$ of the `p`-values for the regression coefficients as tile plot to highlight the identified interaction structure.

```{r trait1_plots_exhaustive}
plotable <- pairs %>%
  pivot_longer(
    cols = starts_with("p_"),
    names_to = "trait",
    names_prefix = "trait ",
    values_to = "p",
    values_drop_na = TRUE
  ) %>%
  mutate(trait = case_when(
    trait == "p_01" ~ "Trait 1",
    trait == "p_02"  ~ "Trait 2"))
tiles <- ggplot(data = plotable, aes(x=var1, y=var2, fill=-log10(p)))+
  geom_tile() +
  facet_wrap(~trait) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_viridis_c()
plot(tiles)
```

The only significant interactions after multiple testing correction are the interaction
between `snp00068` and `snp_00665` as well as `snp_00465` and `snp_00917`.

```{r significant_exhaustive}
pairs %>%
  filter(p_01 < 0.05/15 | p_02 < 0.05/15) %>%
  kable(., linesep = "", digits = 14) %>%
  kable_material(c("striped"))
```

## True epistataic structure

Compare the results of the exhaustive search to the true interaction structure. Notice that the only significant interactions in the exhaustive search are the two with the largest true effects.

```{r simulated_interactions}
true_interactions <- simulated_data$interactions %>%
  mutate(var1 = sprintf(group1, fmt = "snp_%05d")) %>%
  mutate(var2 = sprintf(group2, fmt = "snp_%05d")) %>%
  mutate(trait = case_when(
    trait == 1 ~ "Trait 1",
    trait == 2  ~ "Trait 2")) %>%
  select(-c("group1", "group2"))
X <- true_interactions[, c("var1", "var2")]
X  <- t(apply(X, 1, sort))
true_interactions[,c("var1", "var2")]  <- X

epistatic_pairnames <- comboGeneral(simulated_data$epistatic$name %>% unique(), 2)
true_pairs <- NULL
for (k in seq_len(nrow(epistatic_pairnames))) {
  tib <- dplyr::tibble(var1 = epistatic_pairnames[k, 1],
                        var2 = epistatic_pairnames[k, 2])
  true_pairs <- bind_rows(true_pairs, tib)
}
anti <- anti_join(true_pairs, true_interactions) %>%
  mutate(effect_size = 0)

true_int_plot <- true_interactions %>%
  bind_rows(anti %>% mutate(trait = "Trait 1")) %>%
  bind_rows(anti %>% mutate(trait = "Trait 2"))

true_tiles <- ggplot(data = true_int_plot, aes(x=var1, y=var2, fill=effect_size)) +
  geom_tile() +
  facet_wrap(~trait) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white")
plot(true_tiles)
```

While mvMAPIT does not identify the explicit partner, it still implicates more correct SNPs in this example. All true epistatic SNPs are listed in the following table.

```{r true_interactions}
true_interactions %>%
  kable(., linesep = "") %>%
  kable_material(c("striped"))
```
